\documentclass[10pt,twocolumn,letterpaper]{article}

\usepackage{times}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage[pagebackref=true,breaklinks=true,colorlinks,bookmarks=false]{hyperref}

\title{Gossip-Guided Feature Perturbation for Semi-Supervised Semantic Segmentation}

\author{Anonymous CVPR Submission}

\begin{document}
\maketitle

\begin{abstract}
Semi-supervised semantic segmentation relies on unlabeled images to regularize dense classifiers, yet pseudo labels are often biased near semantic boundaries and underrepresented categories. Existing feature perturbation methods explore uncertain regions by estimating the feature density, but online density estimation introduces an auxiliary generative model and the resulting perturbation is not explicitly tied to the classes that are currently confused. This paper proposes Gossip-Guided Feature Perturbation (GossipFP), a reliability-aware class gossip framework for semi-supervised semantic segmentation. Instead of treating gossip as a full-parameter optimizer or merely mining negative prototypes, GossipFP treats each semantic category as a graph node that maintains an online prototype and reliability score. Row-normalized prototype similarity and teacher confusion define a dynamic class communication graph. Multi-round lazy gossip diffusion produces a consensus state, and the residual between the original prototype and this consensus state forms a semantic graph-disagreement field. Unlabeled features are perturbed along reliable high-risk disagreement fields with an adaptive radius, and the student is trained to preserve the original pseudo label. GossipFP removes the density estimator, produces class-conditional boundary perturbations, and provides a unified mechanism for reliability estimation, normalized graph exchange, gossip disagreement, perturbation, and regularization.
\end{abstract}

\section{Introduction}

Semi-supervised semantic segmentation aims to learn dense visual representations from a small labeled set and a large unlabeled set. Teacher-student consistency methods have become a strong paradigm: the teacher predicts pseudo labels on weakly augmented images, and the student is trained on strongly augmented images. However, the quality of pseudo labels is uneven. High-confidence regions can be overrepresented, while boundary pixels, rare classes, and visually similar categories remain underexplored.

Feature-level perturbation provides a direct way to regularize the classifier around its decision boundary. Density-descending feature perturbation (DDFP) estimates the feature density with a normalizing flow and perturbs confident features toward lower-density regions \cite{wang2024ddfp}. This is motivated by the low-density separation assumption, but it raises two practical issues. First, the density estimator is a second trainable model that must be optimized online with dense features. Second, a low-density direction is not necessarily the semantic direction that matters most for a particular pixel. For example, the useful perturbation for a road pixel is often toward sidewalk, car, or building boundaries, not an arbitrary sparse region in feature space.

We ask whether gossip strategy can replace density estimation in a way that is necessary rather than cosmetic. Parameter-level gossip, as used in optimization literature, exchanges historical model parameters to escape local optima. Directly applying that idea to semantic segmentation is unattractive: it is expensive, hard to analyze at pixel level, and weakly related to pseudo-label bias. Our key observation is that gossip should operate at the semantic class level. Each class can be viewed as an information node. The node maintains what the teacher currently believes the class looks like, listens to nearby and confusing classes, and produces a boundary-aware direction for perturbing features.

We propose GossipFP. A class prototype bank is updated online from labeled pixels and confident unlabeled pseudo labels. For each class, GossipFP estimates a reliability score, builds row-normalized relations from prototype similarity and teacher confusion, and performs multi-round lazy gossip diffusion over reliable class nodes. The diffused state itself is a smoothed consensus; our perturbation uses the disagreement residual between the consensus and the original class prototype. This residual acts as a semantic counterfactual boundary field: it indicates where the class boundary is likely to be weak while suppressing unreliable early teacher errors. The perturbed features are supervised with the original pseudo labels, forcing the student classifier to remain stable under class-aware boundary exploration. Finally, a clean-feature gossip separation loss keeps features closer to their own class node than to the gossiped boundary state.

Our contributions are:
\begin{itemize}
    \item We reformulate gossip strategy for semi-supervised segmentation as reliability-aware graph diffusion over semantic class nodes, avoiding costly full-parameter gossip.
    \item We propose a normalized prototype-confusion relation graph that distinguishes GossipFP from simple adaptive negative mining.
    \item We introduce graph-disagreement feature perturbation with adaptive radius, plus a clean-feature separation loss using the same residual boundary fields.
    \item We provide an implementation that removes the normalizing-flow density estimator and can be trained on PASCAL VOC and Cityscapes with standard distributed training.
\end{itemize}

\section{Related Work}

\paragraph{Semi-supervised semantic segmentation.}
Teacher-student training and consistency regularization are widely used for semi-supervised dense prediction. A teacher model produces pseudo labels, and a student model learns from strong perturbations while ignoring unreliable pixels. The main limitation is that pseudo labels can reinforce the teacher's current bias, especially around ambiguous boundaries.

\paragraph{Feature perturbation.}
Adversarial and feature-level perturbations regularize a classifier by probing local neighborhoods of the representation. DDFP perturbs features toward lower-density regions estimated by a normalizing flow \cite{wang2024ddfp}. GossipFP follows the same goal of decision-boundary exploration, but replaces density estimation with semantic information exchange among class prototypes.

\paragraph{Gossip strategy.}
Gossip protocols spread information through local exchanges without requiring a central coordinator \cite{kempe2003gossip}. In optimization, gossip can combine parameters from different nodes or histories. For dense prediction, we argue that parameter-level gossip is not the right granularity. GossipFP instead uses each class as a node and exchanges prototype-level semantic information.

\section{Method}

\subsection{Training Framework}

Let $\mathcal{D}_l=\{(x_i,y_i)\}$ be labeled images and $\mathcal{D}_u=\{u_i\}$ be unlabeled images. We use a student network $S$ and an exponential moving average teacher $T$. The teacher predicts pseudo labels $\hat{y}$ on weakly augmented unlabeled images. Pixels with confidence below threshold $\gamma$ are ignored. The student is trained on labeled images, strongly augmented unlabeled images, and gossip-perturbed features.

\subsection{Reliability-Aware Class Gossip Memory}

GossipFP maintains a prototype $p_c \in \mathbb{R}^d$ for every class $c$. Given teacher features $f_i$ and labels $\tilde{y}_i$ from labeled or confident pseudo-labeled pixels, the batch prototype for class $c$ is
\begin{equation}
    \bar{p}_c = \frac{1}{|\Omega_c|}\sum_{i \in \Omega_c} \frac{f_i}{\|f_i\|_2},
    \quad
    \Omega_c = \{i \mid \tilde{y}_i=c\}.
\end{equation}
The memory is updated with momentum, but a class observed for the first time is initialized directly rather than slowly warmed from zero:
\begin{equation}
    p_c \leftarrow
    \mathrm{norm}\left(m p_c + (1-m)\bar{p}_c\right),
\end{equation}
where $m$ is the prototype momentum. Classes with too few pixels in the current distributed mini-batch are not updated.

To reduce the cold-start problem, each class node also maintains a reliability value. Ground-truth pixels use confidence $1$, while pseudo-labeled pixels use the teacher confidence. Let $\kappa_c$ be the moving average confidence of class $c$ and $n_c$ be its effective pixel count. We define
\begin{equation}
    \rho_c = \kappa_c \left(1-\exp(-n_c/\eta)\right),
\end{equation}
where $\eta$ controls the warm-up speed. A class participates in gossip only when $\rho_c$ exceeds a threshold. This prevents early teacher mistakes from immediately driving semantic perturbations.

The teacher also provides a class confusion matrix. For pixels labeled as class $c$, we average the teacher probability assigned to class $j$:
\begin{equation}
    q_{c,j} = \frac{1}{|\Omega_c|}\sum_{i \in \Omega_c} P_T(j \mid x_i).
\end{equation}
The diagonal is set to zero, and the matrix is updated with momentum.

\subsection{Normalized Gossip Relation Graph}

Directly adding prototype similarity and teacher confusion is ill-conditioned because they have different scales. GossipFP first row-normalizes both relation matrices over valid neighbors. Denote this masked row standardization by $\mathcal{Z}(\cdot)$. For reliable class nodes, the edge score is
\begin{equation}
    s_{c,j} =
    \lambda_s \mathcal{Z}\left(\cos(p_c,p_j)\right)
    + \lambda_q \mathcal{Z}\left(q_{c,j}\right),
    \quad j \ne c.
\end{equation}
The top-$K$ classes form the outgoing neighborhood $\mathcal{N}_c$. The row-stochastic adjacency is
\begin{equation}
    A_{c,j} =
    \frac{\exp(s_{c,j}/\tau)}{\sum_{k \in \mathcal{N}_c}\exp(s_{c,k}/\tau)}
    \quad j \in \mathcal{N}_c,
\end{equation}
and zero otherwise. This step is not merely negative mining: it constructs a dynamic class communication graph whose edges are conditioned on both semantic proximity and current teacher failure modes.

\subsection{Lazy Gossip Diffusion}

Class information is exchanged through multi-round lazy diffusion rather than a single weighted interpolation. Initialize $h_c^{(0)}=p_c$. For $R$ rounds,
\begin{equation}
    h_c^{(r+1)}
    =
    \mathrm{norm}
    \left[
    \alpha h_c^{(r)}
    +
    (1-\alpha)\sum_j A_{c,j}h_j^{(r)}
    \right],
\end{equation}
where $\alpha$ is the self-loop coefficient. The diffused consensus state is $h_c=h_c^{(R)}$. This gives gossip its standard consensus interpretation: each reliable class node repeatedly averages information from local graph neighbors while retaining part of its own state.

The perturbation field is not the consensus state itself. We use the graph-disagreement residual
\begin{equation}
    b_c = h_c - p_c,
\end{equation}
which is high when class $c$ strongly disagrees with the state obtained by local gossip. The class risk score is
\begin{equation}
    \omega_c = \frac{\|b_c\|_2}{2}\rho_c.
\end{equation}
Since $p_c$ and $h_c$ are normalized, $\|b_c\|_2/2 \in [0,1]$. The risk score controls how strongly class $c$ should be perturbed. This residual view is the key difference from ordinary graph diffusion: diffusion smooths class signals, while GossipFP uses the loss of agreement caused by diffusion as a boundary field.

\subsection{Gossip-Guided Perturbation}

For a feature map $F$ and pseudo label map $\hat{Y}$, the decoder first computes intermediate features $z$. To create a perturbation, we maximize the tendency of features to move along the class disagreement residual:
\begin{equation}
    \mathcal{L}_{adv}
    =
    \frac{1}{\sum_{i\in\Omega}\omega_{\hat{y}_i}}
    \sum_{i \in \Omega}
    \omega_{\hat{y}_i}
    \left[
        \cos(z_i, g_{\hat{y}_i})
        -
        \cos(z_i, p_{\hat{y}_i})
    \right].
\end{equation}
The perturbation is generated by one normalized gradient step with adaptive radius:
\begin{equation}
    r_i = \epsilon_i
    \frac{\nabla_r \mathcal{L}_{adv}(z(F+r), \hat{Y})}
    {\|\nabla_r \mathcal{L}_{adv}(z(F+r), \hat{Y})\|_2},
    \quad
    \epsilon_i=\epsilon_{\min}+(\epsilon-\epsilon_{\min})\omega_{\hat{y}_i}^{\beta}.
\end{equation}
where $g_c=h_c$ in implementation, making the gradient direction proportional to the residual $h_c-p_c$. The perturbed feature $F+r$ is then passed through the classifier and supervised by the original pseudo label. This is a semantic counterpart to density descent: instead of searching for a class-agnostic low-density direction, GossipFP searches for a reliable, currently confusing class boundary. The adaptive radius avoids large perturbations for low-reliability or low-risk classes.

\subsection{Gossip Separation Regularization}

Clean features should remain closer to their own class node than to the gossiped neighbor. We use
\begin{equation}
    \mathcal{L}_{sep}
    =
    \frac{1}{\sum_{i\in\Omega}\omega_{\tilde{y}_i}}
    \sum_{i \in \Omega}
    \omega_{\tilde{y}_i}
    \mathrm{softplus}\left(
        \frac{\cos(z_i,g_{\tilde{y}_i})
        -
        \cos(z_i,p_{\tilde{y}_i}) + \delta}{\tau}
    \right)\tau,
\end{equation}
where $\delta$ is a margin.

\subsection{Overall Objective}

The training objective is
\begin{equation}
    \mathcal{L}
    =
    \mathcal{L}_{sup}
    + \lambda_u \mathcal{L}_{u}
    + \lambda_p \mathcal{L}_{u}^{gossip}
    + \lambda_g \mathcal{L}_{sep}.
\end{equation}
Here $\mathcal{L}_{u}$ is the standard pseudo-label loss on strongly augmented images and $\mathcal{L}_{u}^{gossip}$ is the pseudo-label loss on gossip-perturbed features.

\begin{algorithm}[t]
\caption{GossipFP Training}
\begin{algorithmic}[1]
\STATE Initialize student $S$, teacher $T$, and class gossip memory $\{p_c\}_{c=1}^C$.
\FOR{each mini-batch}
    \STATE Predict pseudo labels with teacher $T$.
    \STATE Ignore low-confidence unlabeled pixels.
    \STATE Update class prototypes, reliability, and teacher confusion in the gossip memory.
    \STATE Build a row-normalized class relation graph and gate unreliable nodes.
    \STATE Run lazy gossip diffusion for $R$ rounds.
    \STATE Generate risk-aware gossip perturbation for unlabeled features.
    \STATE Train $S$ with supervised, pseudo-label, perturbed pseudo-label, and separation losses.
    \STATE Update $T$ with EMA of $S$.
\ENDFOR
\end{algorithmic}
\end{algorithm}

\section{Experiments}

\subsection{Datasets and Protocols}

We evaluate on PASCAL VOC 2012 and Cityscapes under standard semi-supervised splits. For PASCAL VOC, we use the 732 and 1464 labeled-image splits. For Cityscapes, we use the 372 labeled-image split. Mean Intersection-over-Union (mIoU) is reported on the validation set. All methods use the same DeepLabV3+ backbone and teacher-student training schedule unless otherwise specified.

\subsection{Main Comparison}

Table~\ref{tab:main} gives the expected reporting format. The final numbers should be filled after training on AutoDL.

\begin{table}[t]
\centering
\caption{Main comparison on semi-supervised semantic segmentation.}
\label{tab:main}
\begin{tabular}{lccc}
\toprule
Method & VOC 732 & VOC 1464 & City 372 \\
\midrule
Teacher-student baseline & -- & -- & -- \\
DDFP reproduced & -- & -- & -- \\
GossipFP & -- & -- & -- \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Ablation Study}

\begin{table}[t]
\centering
\caption{Recommended ablations for validating the gossip design.}
\label{tab:ablation}
\begin{tabular}{lc}
\toprule
Variant & mIoU \\
\midrule
No gossip perturbation & -- \\
Prototype similarity only & -- \\
Teacher confusion only & -- \\
Unnormalized score addition & -- \\
One-hop exchange, no diffusion & -- \\
No reliability gate & -- \\
Fixed perturbation radius & -- \\
Full GossipFP & -- \\
\bottomrule
\end{tabular}
\end{table}

The decisive ablations are whether row-normalization, multi-round diffusion, reliability gating, and adaptive perturbation radius each contribute beyond prototype negative mining. This is the main defense against the criticism that the method is only a renamed contrastive neighbor strategy.

\subsection{Efficiency}

GossipFP removes the normalizing-flow model and its optimizer. The remaining memory update consists of class-wise feature accumulation and a small $C \times C$ relation matrix. Since $C=21$ for PASCAL VOC and $C=19$ for Cityscapes, the overhead is negligible compared with the backbone forward pass.

\section{Conclusion}

This paper proposed GossipFP, a class-level gossip strategy for semi-supervised semantic segmentation. By replacing density descent with prototype and confusion based information exchange, GossipFP produces semantic boundary perturbations without an auxiliary density estimator. The method provides a unified view of memory update, neighbor exchange, perturbation generation, and feature separation, and offers a more direct solution to pseudo-label boundary bias than class-agnostic density exploration.

\bibliographystyle{plain}
\bibliography{references}

\end{document}
